Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-19 Thesis/03-chap3.qmd
intro to chap 3 revised

Plotting residuals is a standard practice in linear regression diagnostics, essential for identifying deviations from model assumptions such as linearity, homoscedasticity, and normality. Visual inference provides an inferential framework to assess whether residual plots contain patterns inconsistent with model assumptions, typically using a lineup protocol. However, the lineup protocol's reliance on human judgment limits its scalability. This study addresses this limitation by automating the interpretation of residual plots using computer vision models. We develop a distance measure based on Kullback-Leibler divergence to quantify the disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution. We propose a computer vision model to estimate this distance from residual plots, facilitating formal statistical testing and bootstrapping techniques to assess model specification. Our computer vision model shows strong performance, though it performs slightly less effectively on non-linearity visual patterns. The statistical tests based on the estimated distance exhibit lower sensitivity than conventional tests but higher sensitivity than human visual tests. Examples demonstrate the method's effectiveness across different scenarios, highlighting its value in automating the diagnostic process and supplementing traditional methods.

>>>>

Plotting residuals is a recommended practice to diagnose deviations from linear model assumptions such as non-linearity, heteroscedasticity, and non-normality. Using the lineup protocol provides an inferential procedure to statistical assess presence or absence structure in residual plots. As a statistical test, the lineup protocol is less sensitive and applies more broadly than available conventional tests. However, the lineup protocol relies on human judgment which limits its scalability. This work presents a solution by providing a computer vision model to automate the assessment of residual plots. It is trained to predict the distance measure that quantifies the disparity between the residual distribution of a fitted classical normal linear regression model and the reference distribution, based on Kullback-Leibler divergence. From extensive simulation studies, the computer vision model exhibits lower sensitivity than conventional tests but higher sensitivity than human visual tests. It is slightly less effective on non-linearity patterns. Several examples from classical papers and contemporary data illustrate the new procedures, highlighting its usefulness in automating the diagnostic process and supplementing existing methods.
